---
title: "kristen"
output: html_document
---

```{r load-stuff}
library(tidyverse)
library(ggplot2)
```


```{r}
sudors <- read.csv("sudors-data.csv")
cdc <- read.csv("data-table.csv")
head(sudors)

demo_start <- read.csv("SASUMMARY/SASUMMARY__ALL_AREAS_1998_2023.csv")

demo_income <- demo_start %>%
  filter(LineCode == 5) %>% 
  filter(!is.na(Region)) 

demo_disposable <- demo_start %>%
  filter(LineCode == 7) %>%
  filter(!is.na(Region))

# Want a data frame where each column represents a rate, and each row represents a year 

# for each year column 
  # add state name to statename vector and number to income vector 

statenames <- c()
income_values_2020 <- c()
income_values_2021 <- c()
income_values_2022 <- c()
income_values_2023 <- c()

for (i in 1:nrow(demo_income)) {
  statenames <- c(statenames, demo_income[i, "GeoName"])
  income_values_2020 <- c(income_values_2020, demo_income[i, "X2020"])
  income_values_2021 <- c(income_values_2021, demo_income[i, "X2021"])
  income_values_2022 <- c(income_values_2022, demo_income[i, "X2022"])
  income_values_2023 <- c(income_values_2023, demo_income[i, "X2023"])
}
df <- data.frame(statenames, income_values_2020, income_values_2021,income_values_2022, income_values_2023)
df

```


```{r}
table(sudors$Jurisdiction)

sudors <- sudors %>%
  filtered(year == 2020)
```
## Start Coding Different Policy Groups 

### Strat 2 and 3: Drug Checking Equipment
STRATEGY 2: DO NOT SUBJECT FENTANYL TEST STRIPS TO DRUG
PARAPHERNALIA PENALTIES
&
STRATEGY 3: DO NOT SUBJECT DRUG CHECKING EQUIPMENT THAT TESTS
FOR DRUGS OTHER THAN FENTANYL TO DRUG PARAPHERNALIA
PENALTIES
```{r }
# States that follow strategy 2 and 3
states <- c("Alaska", "Arizona", "Colorado", "Connecticut", "Delaware", "District of Columbia", "Georgia", 
            "Hawaii", "Iowa", "Kansas", "Kentucky", "Maine", "Maryland", "Massachusetts", "Michigan", 
            "Minnesota", "Montana", "Nebraska", "Nevada", "New Hampshire", "New Jersey", "New Mexico", 
            "North Carolina", "Ohio", "Oklahoma", "Oregon", "Pennsylvania", "Rhode Island", "Tennessee", 
            "Utah", "Vermont", "Virginia", "West Virginia")

# Define the states supporting access to drug checking equipment (purple states)
support_states <- c("Alaska", "Arizona", "Colorado", "Connecticut", "Delaware", "District of Columbia", "Georgia", 
                    "Minnesota", "Nevada", "New Hampshire", "New Jersey", "New Mexico", "Ohio", "Oklahoma", 
                    "Oregon", "Pennsylvania", "Rhode Island", "Utah", "Vermont", "Virginia", "West Virginia")

# Create a data frame with a 1 for supported states and 0 for others
sudors_policy <- data.frame(
  Jurisdiction = states,
  strat2and3 = ifelse(states %in% support_states, 1, 0)
)


# Merge this with sudors
sudors_policy <- merge(sudors_policy, sudors, by="Jurisdiction")


```

### Strat 4
REQUIRE ACCESS TO MEDICATION FOR ADDICTION
TREATMENT IN STATE OR LOCAL CORRECTIONAL SETTINGS
```{r}

# Define the list of all states
states <- c("Alaska", "Arizona", "Colorado", "Connecticut", "Delaware", "District of Columbia", "Georgia", 
            "Hawaii", "Iowa", "Kansas", "Kentucky", "Maine", "Maryland", "Massachusetts", "Michigan", 
            "Minnesota", "Montana", "Nebraska", "Nevada", "New Hampshire", "New Jersey", "New Mexico", 
            "North Carolina", "Ohio", "Oklahoma", "Oregon", "Pennsylvania", "Rhode Island", "Tennessee", 
            "Utah", "Vermont", "Virginia", "West Virginia")

# Define the states requiring access to medication for addiction treatment (purple states)
treatment_states <- c("Alaska", "Nevada", "Utah", "Colorado", "New Mexico", "Ohio", "Pennsylvania", 
                      "New York", "New Jersey", "Connecticut", "Delaware", "Maryland", "Massachusetts", 
                      "Rhode Island", "Vermont", "New Hampshire")

# Create a data frame with a 1 for treatment states and 0 for others
sudors_policy <- data.frame(
  Jurisdiction = states,
  strat4 = ifelse(states %in% treatment_states, 1, 0)
)

sudors_policy <- merge(sudors_policy, sudors, by="Jurisdiction")

```

### Strat 4 2020
REQUIRE ACCESS TO MEDICATION FOR ADDICTION
TREATMENT IN STATE OR LOCAL CORRECTIONAL SETTINGS
```{r}

# Define the list of all states
states <- c("Alaska", "Arizona", "Colorado", "Connecticut", "Delaware", "District of Columbia", "Georgia", 
            "Hawaii", "Iowa", "Kansas", "Kentucky", "Maine", "Maryland", "Massachusetts", "Michigan", 
            "Minnesota", "Montana", "Nebraska", "Nevada", "New Hampshire", "New Jersey", "New Mexico", 
            "North Carolina", "Ohio", "Oklahoma", "Oregon", "Pennsylvania", "Rhode Island", "Tennessee", 
            "Utah", "Vermont", "Virginia", "West Virginia")

# Define the states requiring access to medication for addiction treatment (purple states)
treatment_states <- c("Alaska", "Nevada", "Utah", "Colorado", "Pennsylvania", 
                   "New Jersey", "Connecticut", "Delaware", "Maryland", "Massachusetts", 
                      "Rhode Island", "Vermont", "New Hampshire")

# Create a data frame with a 1 for treatment states and 0 for others
sudors_policy <- data.frame(
  Jurisdiction = states,
  strat42020 = ifelse(states %in% treatment_states, 1, 0)
)

sudors_policy <- merge(sudors_policy, sudors, by="Jurisdiction")

```

### Strat 7
REQUIRE ALL PUBLIC HIGH SCHOOLS TO STORE NALOXONE
ON SITE FOR RESPONDING TO OVERDOSES AT SCHOOL AND AT
SCHOOL-SPONSORED EVENTS

```{r}
# Define the list of all states
states <- c("Alaska", "Arizona", "Colorado", "Connecticut", "Delaware", "District of Columbia", "Georgia", 
            "Hawaii", "Iowa", "Kansas", "Kentucky", "Maine", "Maryland", "Massachusetts", "Michigan", 
            "Minnesota", "Montana", "Nebraska", "Nevada", "New Hampshire", "New Jersey", "New Mexico", 
            "North Carolina", "Ohio", "Oklahoma", "Oregon", "Pennsylvania", "Rhode Island", "Tennessee", 
            "Utah", "Vermont", "Virginia", "West Virginia")

# Define the states requiring naloxone storage in public high schools (purple states)
naloxone_states <- c("New Jersey", "Maryland", "Rhode Island")

# Create a data frame with a 1 for naloxone states and 0 for others
sudors_policy <- data.frame(
  Jurisdiction = states,
  strat7 = ifelse(states %in% naloxone_states, 1, 0)
)

# Merge with previous data frame sudors
sudors_policy <- merge(sudors_policy, sudors, by = "Jurisdiction")

print(sudors_policy)

```

### Strat 8
INCREASE ACCESS TO SUBSTANCE USE TREATMENT IN
EMERGENCY DEPARTMENT SETTINGS

```{r}
# Define the list of all states
states <- c("Alaska", "Arizona", "Colorado", "Connecticut", "Delaware", "District of Columbia", "Georgia", 
            "Hawaii", "Iowa", "Kansas", "Kentucky", "Maine", "Maryland", "Massachusetts", "Michigan", 
            "Minnesota", "Montana", "Nebraska", "Nevada", "New Hampshire", "New Jersey", "New Mexico", 
            "North Carolina", "Ohio", "Oklahoma", "Oregon", "Pennsylvania", "Rhode Island", "Tennessee", 
            "Utah", "Vermont", "Virginia", "West Virginia")

# Define the states requiring naloxone storage in public high schools (purple states)
naloxone_states <- c("New York", "Massachusetts", "Pennsylvania", "Virginia", "Maryland", "Florida", "Rhode Island")

# Create a data frame with a 1 for naloxone states and 0 for others
sudors_policy <- data.frame(
  Jurisdiction = states,
  strat8 = ifelse(states %in% naloxone_states, 1, 0)
)

# Merge with previous data frame sudors
sudors_policy <- merge(sudors_policy, sudors, by = "Jurisdiction")

print(sudors_policy)

```
### More naxolone distribution
```{r}
# Define the list of all states
states <- c("Alaska", "Arizona", "Colorado", "Connecticut", "Delaware", "District of Columbia", "Georgia", 
            "Hawaii", "Iowa", "Kansas", "Kentucky", "Maine", "Maryland", "Massachusetts", "Michigan", 
            "Minnesota", "Montana", "Nebraska", "Nevada", "New Hampshire", "New Jersey", "New Mexico", 
            "North Carolina", "Ohio", "Oklahoma", "Oregon", "Pennsylvania", "Rhode Island", "Tennessee", 
            "Utah", "Vermont", "Virginia", "West Virginia")

# Define the states requiring naloxone storage in public high schools (purple states)
naloxone_states <- c("Ohio", "Kentucky", "West Virginia", "Virginia", "Maryland", "Delaware", "New Jersey", "Oregon", "Arizona", "New Mexico", "Tennessee")

# Create a data frame with a 1 for naloxone states and 0 for others
sudors_policy <- data.frame(
  Jurisdiction = states,
  stratnax = ifelse(states %in% naloxone_states, 1, 0)
)

# Merge with previous data frame sudors
sudors_policy <- merge(sudors_policy, sudors, by = "Jurisdiction")

print(sudors_policy)

```

### Decriminalization
```{r}
# Define the list of all states
states <- c("Alaska", "Arizona", "Colorado", "Connecticut", "Delaware", "District of Columbia", "Georgia", 
            "Hawaii", "Iowa", "Kansas", "Kentucky", "Maine", "Maryland", "Massachusetts", "Michigan", 
            "Minnesota", "Montana", "Nebraska", "Nevada", "New Hampshire", "New Jersey", "New Mexico", 
            "North Carolina", "Ohio", "Oklahoma", "Oregon", "Pennsylvania", "Rhode Island", "Tennessee", 
            "Utah", "Vermont", "Virginia", "West Virginia")

# Define the states requiring naloxone storage in public high schools (purple states)
naloxone_states <- c("Oregon", "Kansas", "Texas", "Kentucky", "Maryland", "New York", "Vermont", "New Hampshire", "Massachusetts", "Rhode Island", "Maine")

# Create a data frame with a 1 for naloxone states and 0 for others
sudors_policy <- data.frame(
  Jurisdiction = states,
  stratdecrim = ifelse(states %in% naloxone_states, 1, 0)
)

# Merge with previous data frame sudors
sudors_policy <- merge(sudors_policy, sudors, by = "Jurisdiction")

print(sudors_policy)

```

```{r}
# Define the list of all states
states <- c("AL", "AK", "AZ", "AR", "CA", "CO", "CT", "DE", "FL", "GA", "HI", "ID", "IL", "IN", "IA", "KS", "KY", "LA", "ME", "MD", "MA", "MI", "MN", "MS", "MO", "MT", "NE", "NV", "NH", "NJ", "NM", "NY", "NC", "ND", "OH", "OK", "OR", "PA", "RI", "SC", "SD", "TN", "TX", "UT", "VT", "VA", "WA", "WV", "WI", "WY")

# Define the states requiring naloxone storage in public high schools (purple states)
naloxone_states <- c("AK", "CA", "AZ", "NM", "CO", "TX", "NE", "SD", "IA", "WI", "IL", "TN", "WV", "GA", "FL", "SC", "NH")

# Create a data frame with a 1 for naloxone states and 0 for others
cdc_policy <- data.frame(
  STATE = states,
  stratdecrim = ifelse(states %in% naloxone_states, 1, 0)
)

# Merge with previous data frame sudors
cdc_policy <- merge(cdc_policy, cdc, by = "STATE")
```


## Hypothesis testing
#### Those who live in high drug rate groups 
```{r}
high <- sudors %>%
  filter(Jurisdiction == "Indiana" | Jurisdiction == "Maine" | Jurisdiction == "Delaware" | Jurisdiction == "Oregon" | Jurisdiction == "Kentucky")

low <- sudors %>%
  filter(Jurisdiction == "Nebraska" | Jurisdiction == "Florida" | Jurisdiction == "Utah" | Jurisdiction == "Hawaii")

test <- t.test(high$alldrug_deaths, low$alldrug_deaths, alternative = 'greater')
test
```

### Those who live in states that do not support strategies 2 and 1 die more from opiod overdoses
```{r}

```

### Example hypothesis: Those who live in states that do not support strategies 2 and 1 use more opiods
```{r}
support <- df_2and3 %>%
  filter(strat2and3 == 1)

no_support <- df_2and3 %>%
  filter(strat2and3 == 0)

test <- t.test(no_support$alldrug_rate, support$alldrug_rate, alternative = 'greater')
test
```

### Those who live in states that support strategy 4 die less from all drugs
```{r}

support <- sudors_policy %>%
  filter(strat4 == 1)

no_support <- sudors_policy %>%
  filter(strat4 == 0)

test <- t.test(support$alldrug_rate, no_support$alldrug_rate, alternative = 'less')
test
```

```{r}

support <- sudors_policy %>%
  filter(strat4 == 1)

no_support <- sudors_policy %>%
  filter(strat4 == 0)

test <- t.test(no_support$black_nh_rate, support$black_nh_rate, alternative = 'greater')
test
```

## Hypothesis: States that do not support Naxalone in public schools have more deaths in persons under the age of 15

```{r}
support <- sudors_policy %>%
  filter(strat7 == 1) 

no_support <- sudors_policy %>%
  filter(strat7 == 0)

test <- t.test(no_support$heroin_percent, support$her, alternative = 'greater')
test
```
```{r}
support <- sudors_policy %>%
  filter(strat7 == 1) 

no_support <- sudors_policy %>%
  filter(strat7 == 0)

test <- t.test(support$age_under15_deaths, no_support$age_under15_deaths, alternative = 'less')
test
```

###

```{r}
support <- sudors_policy %>%
  filter(strat8 == 1) 

no_support <- sudors_policy %>%
  filter(strat8 == 0)

test <- t.test(support$hx_anydrug_percent, no_support$hx_anydrug_percent, alternative = 'less')
test
```

```{r}
support <- sudors_policy %>%
  filter(strat8 == 1) 

no_support <- sudors_policy %>%
  filter(strat8 == 0)

test <- t.test(no_support$hx_anydrug_percent, support$hx_anydrug_percent, alternative = 'greater')
test
```

### HYpothesis: There is a difference in states that utilize policy 8 and those who do not
```{r}

# Test the hypothesis here
anova <- aov(alldrug_rate ~ strat4, data = sudors_policy) # Dependent variable ~ independent variable
summary(anova)
```


```{r}
support <- sudors_policy %>%
  filter(stratnax == 1) 

no_support <- sudors_policy %>%
  filter(stratnax == 0)

test <- t.test(no_support$heroin_rate, support$heroin_rate, alternative = 'greater')
test
```

```{r}
support <- sudors_policy %>%
  filter(stratnax == 1) 

no_support <- sudors_policy %>%
  filter(stratnax == 0)

test <- t.test(support$opioids_nostim_percent, no_support$opioids_nostim_percent, alternative = 'less')
test
```

## Linear Modeling 
### 

```{r fig.height=10, fig.width=12}
# SIGNIFICANT


#m <- lm(male_rate~factor(stratdecrim), data = sudors_policy)
#summary(m)

#m <- lm(heroin_rate~stratdecrim, data = sudors_policy)
#summary(m)

#NOT
#m <- lm(imfs_rate~stratdecrim, data = sudors_policy)
#summary(m)

#m <- lm(opioids_rate~stratdecrim, data = sudors_policy)
#summary(m)


# <- lm(cocaine_rate~stratdecrim, data = sudors_policy)
#summary(m)

#m <- lm(illegal_drugs_deaths~stratdecrim, data = sudors_policy)
#summary(m)


#m <- lm(RATE ~ stratdecrim, data = cdc_policy)
#summary(m)

#ggplot(cdc, aes(YEAR, RATE)) +
 # geom_point() + 
  #geom_smooth(method='lm') + facet_wrap(~STATE) + geom_vline(xintercept = 2020)

cdc$YEAR <- as.numeric(cdc$YEAR)

cdc_2020 <- cdc %>%
  filter(YEAR < 2020)

coeffs <- cdc_2020 %>%
  group_by(STATE, stratdecrim) %>%
  do({
    model <- lm(RATE ~ YEAR, data = .)
    data.frame(Intercept = coef(model)[1], Slope = coef(model)[2])
  })

coeffs <- coeffs %>%
  arrange(Slope)

print(coeffs)


```