---
title: "Main"
output: html_document
---

## Load Datasets and Libraries

```{r load-stuff}
library(tidyverse)
library(ggplot2)
```

```{r pre-process}
sudors <- read.csv("sudors-data.csv")
cdc <- read.csv('data-table.csv')

```

## Processing


### Policy 1: Sudors
```{r policy1-SUDORS}

# Define the list of all states
states <- c("Alaska", "Arizona", "Colorado", "Connecticut", "Delaware", "District of Columbia", "Georgia", 
            "Hawaii", "Iowa", "Kansas", "Kentucky", "Maine", "Maryland", "Massachusetts", "Michigan", 
            "Minnesota", "Montana", "Nebraska", "Nevada", "New Hampshire", "New Jersey", "New Mexico", 
            "North Carolina", "Ohio", "Oklahoma", "Oregon", "Pennsylvania", "Rhode Island", "Tennessee", 
            "Utah", "Vermont", "Virginia", "West Virginia")

# POLICY 1
treatment_states <- c("Colorado", "Connecticut", "Delaware", "Georgia", "Hawaii", "Maine", "Maryland", "Montana", "Nevada", "New Hampshire", "New Jersey", "New Mexico", "Ohio", "Oregon", "Tennessee", "Vermont", "Virginia", "West Virginia")

# Create a data frame with a 1 for treatment states and 0 for others
sudors_policy1 <- data.frame(
  Jurisdiction = states,
  strat1 = ifelse(states %in% treatment_states, 1, 0)
)

sudors_policy1 <- merge(sudors_policy1, sudors, by="Jurisdiction")

```

### Policy 1: NCHS
```{r policy1-NCHS}

# Define the list of all states
states <- c("AL", "AK", "AZ", "AR", "CA", "CO", "CT", "DE", "FL", "GA", "HI", "ID", "IL", "IN", "IA", "KS", "KY", "LA", "ME", "MD", "MA", "MI", "MN", "MS", "MO", "MT", "NE", "NV", "NH", "NJ", "NM", "NY", "NC", "ND", "OH", "OK", "OR", "PA", "RI", "SC", "SD", "TN", "TX", "UT", "VT", "VA", "WA", "WV", "WI", "WY")

# POLICY 1
treatment_states <- c("ME", "VT", "NH", "CT", "NJ", "DE", "MD", "DC", "MD", "VA", "WV", "OH", "IN", "IL", "VA", "TN", "GA", "FL", "MO", "LA", "AR", "MT", "SD", "CO", "NM", "CA", "NV", "OR")

# Create a data frame with a 1 for treatment states and 0 for others
nchs_policy1 <- data.frame(
  STATE = states,
  strat1 = ifelse(states %in% treatment_states, 1, 0)
)

nchs_policy1 <- merge(nchs_policy1, cdc, by="STATE")
```


### Policy 2: Sudors
```{r policy2-SUDORS}

# Define the list of all states
states <- c("Alaska", "Arizona", "Colorado", "Connecticut", "Delaware", "District of Columbia", "Georgia", 
            "Hawaii", "Iowa", "Kansas", "Kentucky", "Maine", "Maryland", "Massachusetts", "Michigan", 
            "Minnesota", "Montana", "Nebraska", "Nevada", "New Hampshire", "New Jersey", "New Mexico", 
            "North Carolina", "Ohio", "Oklahoma", "Oregon", "Pennsylvania", "Rhode Island", "Tennessee", 
            "Utah", "Vermont", "Virginia", "West Virginia")

# Define the states requiring access to medication for addiction treatment (purple states)
treatment_states <- c("Alaska", "Nevada", "Utah", "Colorado", "New Mexico", "Ohio", "Pennsylvania", 
                      "New York", "New Jersey", "Connecticut", "Delaware", "Maryland", "Massachusetts", 
                      "Rhode Island", "Vermont", "New Hampshire")

# Create a data frame with a 1 for treatment states and 0 for others
sudors_policy2 <- data.frame(
  Jurisdiction = states,
  strat4 = ifelse(states %in% treatment_states, 1, 0)
)

sudors_policy2 <- merge(sudors_policy2, sudors, by="Jurisdiction")

```

### Policy 3: NCHS
```{r}
# Define the list of all states
states <- c("AL", "AK", "AZ", "AR", "CA", "CO", "CT", "DE", "FL", "GA", "HI", "ID", "IL", "IN", "IA", "KS", "KY", "LA", "ME", "MD", "MA", "MI", "MN", "MS", "MO", "MT", "NE", "NV", "NH", "NJ", "NM", "NY", "NC", "ND", "OH", "OK", "OR", "PA", "RI", "SC", "SD", "TN", "TX", "UT", "VT", "VA", "WA", "WV", "WI", "WY")

decrim_states <- c("AK", "CA", "AZ", "NM", "CO", "TX", "NE", "SD", "IA", "WI", "IL", "TN", "WV", "GA", "FL", "SC", "NH")

cdc_policy <- data.frame(
  STATE = states,
  stratdecrim = ifelse(states %in% decrim_states, 1, 0)
)

# Merge with previous data frame sudors
cdc_policy <- merge(cdc_policy, cdc, by = "STATE")

```


### Policy3 SUDORS
```{r}
# Define the list of all states
states <- c("Alaska", "Arizona", "Colorado", "Connecticut", "Delaware", "District of Columbia", "Georgia", 
            "Hawaii", "Iowa", "Kansas", "Kentucky", "Maine", "Maryland", "Massachusetts", "Michigan", 
            "Minnesota", "Montana", "Nebraska", "Nevada", "New Hampshire", "New Jersey", "New Mexico", 
            "North Carolina", "Ohio", "Oklahoma", "Oregon", "Pennsylvania", "Rhode Island", "Tennessee", 
            "Utah", "Vermont", "Virginia", "West Virginia")

naloxone_states <- c("Oregon", "Kansas", "Texas", "Kentucky", "Maryland", "New York", "Vermont", "New Hampshire", "Massachusetts", "Rhode Island", "Maine")

# Create a data frame with a 1 for naloxone states and 0 for others
sudors_policy3 <- data.frame(
  Jurisdiction = states,
  stratdecrim = ifelse(states %in% naloxone_states, 1, 0)
)

# Merge with previous data frame sudors
sudors_policy3 <- merge(sudors_policy3, sudors, by = "Jurisdiction")

```
## Correlational Tests

### Hypothesis 1
States who support policy 1 have less average drug rates than other states over the time period in the datasets

```{r hypothesis1}
support <- sudors_policy1 %>%
  filter(strat1 == 1)

no_support <- sudors_policy1 %>%
  filter(strat1 == 0)

test <- t.test(support$alldrug_rate, no_support$alldrug_rate, alternative = 'less')
test
```
### Hypothesis 2

### Hypothesis 1
States who support policy 1 have less average drug rates than other states over the time period in the datasets

```{r hypothesi1}
support <- nchs_policy1 %>%
  filter(strat1 == 1)

no_support <- nchs_policy1 %>%
  filter(strat1 == 0)

test <- t.test(support$RATE, no_support$RATE, alternative = 'less')
test
```
```{r hypothesis1-SUDORS}
#support <- sudors_policy1 %>%
  #filter(strat1 == 1)

#no_support <- sudors_policy1 %>%
 # filter(strat1 == 0)

#test <- t.test(support$RATE, no_support$RATE, alternative = 'less')
#test
```

### Hypothesis 2
```{r}

support <- sudors_policy2 %>%
  filter(strat4 == 1)

no_support <- sudors_policy2 %>%
  filter(strat4 == 0)

test <- t.test(support$age_15_24_rate, no_support$age_15_24_rate, alternative = 'less')
test
```

### Hypothesis 3
```{r}

support <- sudors_policy3 %>%
  filter(stratdecrim == 1)

no_support <- sudors_policy3 %>%
  filter(stratdecrim == 0)

test <- t.test(support$age_15_24_rate, no_support$age_15_24_rate, alternative = 'less')
test
```


## Linear Modeling 
```{r fig.height=10, fig.width=12}
# SIGNIFICANT
cdc_policy$YEAR <- as.numeric(cdc_policy$YEAR)

cdc_2020 <- cdc_policy %>%
  filter(YEAR < 2020)

# Not including until after 2020
coeffs <- cdc_2020 %>%
  group_by(STATE, stratdecrim) %>%
  do({
    model <- lm(RATE ~ YEAR, data = .)
    data.frame(Intercept = coef(model)[1], Slope = coef(model)[2])
  })

coeffs <- coeffs %>%
  arrange(Slope)

print(coeffs)

coeffs_one <- coeffs %>%
  filter(stratdecrim == 1)
  
mean(coeffs_one$Slope)

coeffs_two <- coeffs %>%
  filter(stratdecrim == 0)
  
mean(coeffs_two$Slope, na.rm = TRUE)









# Including after 2020
coeffs <- cdc_policy %>%
  group_by(STATE, stratdecrim) %>%
  do({
    model <- lm(RATE ~ YEAR, data = .)
    data.frame(Intercept = coef(model)[1], Slope = coef(model)[2])
  })

coeffs <- coeffs %>%
  arrange(Slope)

print(coeffs)

coeffs_one <- coeffs %>%
  filter(stratdecrim == 1)
  
mean(coeffs_one$Slope)

coeffs_two <- coeffs %>%
  filter(stratdecrim == 0)
  
mean(coeffs_two$Slope, na.rm = TRUE)

# Pvalue
library(dplyr)

coeffs <- cdc %>%
  group_by(STATE) %>%
  do({
    model <- lm(RATE ~ YEAR, data = .)
    summary_model <- summary(model)
    data.frame(
      Intercept = coef(model)[1],
      Slope = coef(model)[2],
      R_Squared = summary_model$r.squared,
      P_Value = summary_model$coefficients[, 4]  # Extract p-value for Slope
    )
  })

print(coeffs)

ggplot(cdc, aes(YEAR, RATE)) +
  geom_point() + 
  geom_smooth(method='lm') + facet_wrap(~STATE) + geom_vline(xintercept = 2020)


```